Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each $\tau \in \ssorts{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in $\I$,
and each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$ 
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where 
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is 
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$;
otherwise, it is \emph{$T$-unsatisfiable}.
A formula $\varphi$ entails $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Formulas $\varphi$ and $\psi$ are \emph{equisatisfiable} if
$\varphi$ is $T$-satisfiable if and only if $\psi$ is $T$-satisfiable.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the type and function symbols in this theory as \emph{uninterpreted}.

@@

Let us fix a signature $\Sigma$, a $\Sigma$-formula
$\varphi$ in definitional form with respect to $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature
$\extendsig{ \Sigma }$ and a $\extendsig{ \Sigma }$-formula $\psi$ such that
$\varphi$ and $\psi$ are equisatisfiable.
\jb{Equisatisfiability as defined in Sect.\ 2 requires the same $\Sigma$ and
the same $T$ for both formulas, which isn't the case here.}
The intuition behind this translation
is to use an uninterpreted type \ct{to be defined} to abstract the set of
\relax{relevant} tuples for each recursive function $\con{f}$ and restrict the
quantification of the function definition for $\con{f}$ to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \mathrel{\cup} \Sigma_{\euf}$, where
$\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \sortint \rightarrow \sortint },\, \const{c} : { \sortint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} n : {\sortint}.\;\, \lite\bigl( n \leq 0,\; \con{s}(n) \teq 0,\;
  \con{s}( n ) \teq n + \con{s}( n - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation} 
%
Here, $\const{s}$ is a function that returns the sum of the positive integers
up to $n$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma_u}$ and states that the sum of all
positive numbers between $0$ and $\const{c}$ is greater than $100$. This formula is
satisfiable with a model that interprets $\const{c}$ as an integer greater than or
equal to $14$. Due to the universal quantifier,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{\con{s}}$ and the uninterpreted function $\farg{\con{s}}$ of type $\fargsort{\con{s}}
\rightarrow \sortint$. The result of $\conv( \varphi,\, \ppos )$ is the pair $(
\psi,\, \ltrue )$, where (after simplification) $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[c]
  & \phantom{{\wedge}\; }\bigl(
    \!\begin{aligned}[t]
      \forall a : \fargsort{\con{s}}.\; & \phantom{{\wedge}\;}\lite\bigl( \farg{\con{s}}( a ) \leq 0,\;
        \con{s}(\farg{\con{s}}( a )) \teq 0,\;
        \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 ) \bigr)
\\[-\jot]
      & {\wedge}\; \exists a : {\fargsort{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \farg{\con{s}}( a )-1 \bigr)
\end{aligned}
\\[-\jot]
 & {\wedge}\; \bigl( \con{s}( c ) > 100 \wedge \exists a : {\fargsort{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \con{c} \bigr)
\end{aligned}
\end{equation}
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{\con{s}}$, thus enabling the finite model finding
techniques by Reynolds et al.\ \cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{\con{s}}$. Moreover, all occurrences of $a$ in this formula are beneath
applications of the uninterpreted function $\farg{\con{s}}$, implying that the
formula is in the essentially uninterpreted fragment,
for which Ge and de Moura \cite{GeDeM-CAV-09} provide a complete instantiation
procedure. We observe that both
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.\xend
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{\con{s}}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{equation} \label{eq:approx-interp}
\lambda x\typ{Int}. \lite\bigl( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )\bigr)
\end{equation}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g.\ $\con{s}( -1 )$ and $\con{s}( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are admissible, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $\con{s}( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp.\ nonexistence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For example, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is not an issue, because
$s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

>>>>

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each $\tau \in \ssorts{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in $\I$,
and each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$ 
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where 
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is 
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$;
otherwise, it is \emph{$T$-unsatisfiable}.
A formula $\varphi$ entails $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the type and function symbols in this theory as \emph{uninterpreted}.

@@

Let us fix a signature $\Sigma$, a $\Sigma$-formula
$\varphi$ in definitional form with respect to $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature
$\extendsig{ \Sigma }$ and a $\extendsig{ \Sigma }$-formula $\psi$ such that
$\psi$ is satisfiable if and only if $\varphi$ is satisfiable---i.e.,
$\varphi$ and $\psi$ are \emph{equisatisfiable}.
The intuition behind this translation
is to use an uninterpreted type \ct{to be defined} to abstract the set of
\relax{relevant} tuples for each recursive function $\con{f}$ and restrict the
quantification of the function definition for $\con{f}$ to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \mathrel{\cup} \Sigma_{\euf}$, where
$\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \sortint \rightarrow \sortint },\, \const{c} : { \sortint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} n : {\sortint}.\;\, \lite\bigl( n \leq 0,\; \con{s}(n) \teq 0,\;
  \con{s}( n ) \teq n + \con{s}( n - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation} 
%
Here, $\const{s}$ is a function that returns the sum of the positive integers
up to $n$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma_u}$ and states that the sum of all
positive numbers between $0$ and $\const{c}$ is greater than $100$. This formula is
satisfiable with a model that interprets $\const{c}$ as an integer greater than or
equal to $14$. Due to the universal quantifier,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{\con{s}}$ and the uninterpreted function $\farg{\con{s}}$ of type $\fargsort{\con{s}}
\rightarrow \sortint$. The result of $\conv( \varphi,\, \ppos )$ is the pair $(
\psi,\, \ltrue )$, where (after simplification) $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[c]
  & \phantom{{\wedge}\; }\bigl(
    \!\begin{aligned}[t]
      \forall a : \fargsort{\con{s}}.\; & \phantom{{\wedge}\;}\lite\bigl( \farg{\con{s}}( a ) \leq 0,\;
        \con{s}(\farg{\con{s}}( a )) \teq 0,\;
        \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 ) \bigr)
\\[-\jot]
      & {\wedge}\; \exists a : {\fargsort{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \farg{\con{s}}( a )-1 \bigr)
\end{aligned}
\\[-\jot]
 & {\wedge}\; \bigl( \con{s}( c ) > 100 \wedge \exists a : {\fargsort{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \con{c} \bigr)
\end{aligned}
\end{equation}
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{\con{s}}$, thus enabling the finite model finding
techniques by Reynolds et al.\ \cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{\con{s}}$. Moreover, all occurrences of $a$ in this formula are beneath
applications of the uninterpreted function $\farg{\con{s}}$, implying that the
formula is in the essentially uninterpreted fragment,
for which Ge and de Moura \cite{GeDeM-CAV-09} provide a complete instantiation
procedure. Both
%%% Added "apparently". Or are we sure?
\cvc and \ziii apparently run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) within 100 milliseconds.\xend
\end{example}

The translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{\con{s}}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$\con{c}$ as $14$, 
and $\con{s}$ as the (almost constant) function:
%
%% TYPESETTING: \mathbin and to make it fit nicely
\begin{equation} \label{eq:approx-interp}
\lambda x : \sortint.\,\, \lite( x \mathbin{\teq} 0,\, 0,\,
  \lite( x \mathbin{\teq} 1,\, 1,\,
    \lite( x \mathbin{\teq} 2,\, 3,\,
      \lite(\ldots,\; \lite( x \mathbin{\teq} 13,\, 91,\, 105 )\negvthinspace \ldots ))))
\end{equation}
%
In other words, $\const{s}$ is interpreted as a function mapping $x$ to the sum
of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$, and $105$
otherwise.
The interpretation $\I$, restricted to $\Sigma$, is {not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ wrongly interprets $\con{s}( x )$ as $105$ when $x < 0$ or $x > 14$.

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, we claim that $\conv(\varphi)$ is
equisatisfiable to $\varphi$ for any input $\varphi$. The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv(
\varphi )$ implies the existence (resp.\ nonexistence) of a model for
$\varphi$. Moreover, information in models for $\conv( \varphi )$ contains
relevant information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$,
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$
as $14$. In fact, we claim that there exists a model of $\varphi$ that
coincides with each model of $\conv( \varphi )$ on its interpretation of all
functions in $\sfuns{\Sigma}$ other than those in $\sfundefs{\Sigma}$, i.e. $\{
s \}$ in the previous example. From a practical point of view, this is not an
issue, because $s$ is the very function that was explicitly defined by the
user, and hence already has an intended interpretation.
